---
title: "HR_script"
author: "crodrigu"
date: "February 05, 2017"
output: html_document
---
<hr>\clearpage

#The Business Questions

The data set being used contains information regarding the employment statsitics and habits of a company from the HR department.  The following are the criteria illustrated in the dataset:

1. Satisfaction level - a number between 0 and 1 - level of satisfaction in current function

2. Evaluation - number between 0 and 1 - 

3. Number of projects - number of projects

4. Average monthly hours

5. Time spent company - number of years worked for the company

6. Work accident - 0 if you've had a work accident, 1 if you have

7. Left - 1 if you've left the company, 0 means you haven't

8. Promotion - 1 if you've had a promotion in the last 5 years, and 0 and if you haven't

9. Department - qualitative (string) 

10. Salary - text - "high", "medium" "low"

Based on the given criteria the business questions that were investigated for the purpose of report: 

1. Why are employees leaving? 

2. What are the three most important factors for people leaving as compared to staying? 
a) Are we overworking the top performers? 

b) Is money most important? Is satsifaction most important? 

c) Department vs. Global Trends - do we have bad managers? 

3. How can we keep employees using the information that we derive from the data? Can we identify people who are leaving

4. Can we identify loyal employees? 

5. Turnover per department? 

```{r setuplibraries, echo=FALSE, message=FALSE}
suppressWarnings(source("AnalyticsLibraries/library.R"))
# Package options
suppressWarnings(ggthemr('fresh'))  # ggplot theme
opts_knit$set(progress=FALSE, verbose=FALSE)
opts_chunk$set(echo=FALSE, fig.align="center", fig.width=10, fig.height=6.35, results="asis")
options(knitr.kable.NA = '')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("read_data.R")
mean(emp_data[,1])
```



```{r setupdata1E, echo=FALSE, tidy=TRUE}
# Please ENTER the name of the file with the data used. The file should be a .csv with one row per observation (e.g. person) and one column per attribute. Do not add .csv at the end, make sure the data are numeric.
datafile_name = "data/HR_comma_sep.csv"

# Please enter the minimum number below which you would like not to print - this makes the readability of the tables easier. Default values are either 10e6 (to print everything) or 0.5. Try both to see the difference.
MIN_VALUE = 0.5

# Please enter the maximum number of observations to show in the report and slides. 
# DEFAULT is 10. If the number is large the report may be slow.
max_data_report = 10

```

## Part 1

Number of current employees:
```{r echo=FALSE}
hr_ds_current <- emp_data %>% filter(left == 0)
(tot_current <- nrow(hr_ds_current))

```

Number of employees who have left:
```{r echo=FALSE}
hr_ds_left <- emp_data %>% filter(left == 1)
(tot_left <- nrow(hr_ds_left))
```

 

## Part 2: Comparison between current employees and those who left:

*Employee Satisfaction*

```{r echo=FALSE, tidy=TRUE}
boxplot(hr_ds_current$satisfaction_level, hr_ds_left$satisfaction_level, names = c("current", "left"))
employee_satisfaction<-t.test(hr_ds_current$satisfaction_level, hr_ds_left$satisfaction_level, var.equal = FALSE)
```


The above test and boxplot shows that the average satisfaction level of  employees who left is lower than those who are still working at the company.

*Employee Evaluation*

```{r echo=FALSE}
boxplot(hr_ds_current$last_evaluation, hr_ds_left$last_evaluation, names = c("current", "left"))
employee_evaluation<-t.test(hr_ds_current$last_evaluation, hr_ds_left$last_evaluation, var.equal = FALSE)
```


The above test p-value and boxplot shows that there is no significant difference in mean associated to the last evaluation between current and employees who left.  In other words, the evaluation that's given is not a reason that employees leave either. 

*Employee Working Hours*

```{r echo=FALSE}
boxplot(hr_ds_current$average_montly_hours, hr_ds_left$average_montly_hours, names = c("current", "left"))
employee_hours<-t.test(hr_ds_current$average_montly_hours, hr_ds_left$average_montly_hours, var.equal = FALSE)

```

The above statistic calculation shows that the average working hours of employees who left are significantly higher than the current ones. This could be identified as one of the reasons that the employees left.

*General Term of Employees*

```{r echo=FALSE}
boxplot(hr_ds_current$time_spend_company, hr_ds_left$time_spend_company, names = c("current", "left"))
employee_spend<-t.test(hr_ds_current$time_spend_company, hr_ds_left$time_spend_company, var.equal = FALSE)
```

The current employees have spent on average fewer years than the employees who left.  This could indicate that the number of years that the employees who left worked is the general term of an employee at the company. This can be the natural turnover rate.

*Number of Employee Projects*

```{r echo=FALSE}
boxplot(hr_ds_current$number_project, hr_ds_left$number_project, names = c("current", "left"))
num_proj<-t.test(hr_ds_current$number_project, hr_ds_left$number_project, var.equal = FALSE)
```

The average number of projects that employees who leave is higher than that of those who are still employed at the company.

#Department specific analysis

In order to analyse the data per department, we create a table per department.
```{r echo=FALSE}

accounting <- emp_data[which(emp_data$Department=="accounting"),]

hr <- emp_data[which(emp_data$Department=="hr"),]

IT<-emp_data[which(emp_data$Department=="IT"),]

management <- emp_data[which(emp_data$Department=="management"),]

product_mng <- emp_data[which(emp_data$Department=="product_mng"),]

marketing <- emp_data[which(emp_data$Department=="marketing"),]

RandD <- emp_data[which(emp_data$Department=="RandD"),]

sales <- emp_data[which(emp_data$Department=="sales"),]

support <- emp_data[which(emp_data$Department=="support"),]

technical <- emp_data[which(emp_data$Department=="technical"),]
```


Then, we calculate the mean parameters of each department.

```{r echo=FALSE}
accounting_mean<-c(round(mean(accounting[,1]),2),round(mean(accounting[,2]),2),round(mean(accounting[,3]),2),round(mean(accounting[,4]),2),round(mean(accounting[,5]),2),round(mean(accounting[,6]),2),round(mean(accounting[,7]),2),round(mean(accounting[,8]),2),round((sum(accounting$salary=="low")+sum(accounting$salary=="medium")*2+sum(accounting$salary=="high")*3)/nrow(accounting),2))

hr_mean<-c(round(mean(hr[,1]),2),round(mean(hr[,2]),2),round(mean(hr[,3]),2),round(mean(hr[,4]),2),round(mean(hr[,5]),2),round(mean(hr[,6]),2),round(mean(hr[,7]),2),round(mean(hr[,8]),2),round((sum(hr$salary=="low")+sum(hr$salary=="medium")*2+sum(hr$salary=="high")*3)/nrow(hr),2))

IT_mean<-c(round(mean(IT[,1]),2),round(mean(IT[,2]),2),round(mean(IT[,3]),2),round(mean(IT[,4]),2),round(mean(IT[,5]),2),round(mean(IT[,6]),2),round(mean(IT[,7]),2),round(mean(IT[,8]),2),round((sum(IT$salary=="low")+sum(IT$salary=="medium")*2+sum(IT$salary=="high")*3)/nrow(IT),2))

management_mean<-c(round(mean(management[,1]),2),round(mean(management[,2]),2),round(mean(management[,3]),2),round(mean(management[,4]),2),round(mean(management[,5]),2),round(mean(management[,6]),2),round(mean(management[,7]),2),round(mean(management[,8]),2),round((sum(management$salary=="low")+sum(management$salary=="medium")*2+sum(management$salary=="high")*3)/nrow(management),2))

marketing_mean<-c(round(mean(marketing[,1]),2),round(mean(marketing[,2]),2),round(mean(marketing[,3]),2),round(mean(marketing[,4]),2),round(mean(marketing[,5]),2),round(mean(marketing[,6]),2),round(mean(marketing[,7]),2),round(mean(marketing[,8]),2),round((sum(marketing$salary=="low")+sum(marketing$salary=="medium")*2+sum(marketing$salary=="high")*3)/nrow(marketing),2))

product_mng_mean<-c(round(mean(product_mng[,1]),2),round(mean(product_mng[,2]),2),round(mean(product_mng[,3]),2),round(mean(product_mng[,4]),2),round(mean(product_mng[,5]),2),round(mean(product_mng[,6]),2),round(mean(product_mng[,7]),2),round(mean(product_mng[,8]),2),round((sum(product_mng$salary=="low")+sum(product_mng$salary=="medium")*2+sum(product_mng$salary=="high")*3)/nrow(product_mng),2))

RandD_mean<-c(round(mean(RandD[,1]),2),round(mean(RandD[,2]),2),round(mean(RandD[,3]),2),round(mean(RandD[,4]),2),round(mean(RandD[,5]),2),round(mean(RandD[,6]),2),round(mean(RandD[,7]),2),round(mean(RandD[,8]),2),round((sum(RandD$salary=="low")+sum(RandD$salary=="medium")*2+sum(RandD$salary=="high")*3)/nrow(RandD),2))

sales_mean<-c(round(mean(sales[,1]),2),round(mean(sales[,2]),2),round(mean(sales[,3]),2),round(mean(sales[,4]),2),round(mean(sales[,5]),2),round(mean(sales[,6]),2),round(mean(sales[,7]),2),round(mean(sales[,8]),2),round((sum(sales$salary=="low")+sum(sales$salary=="medium")*2+sum(sales$salary=="high")*3)/nrow(sales),2))

support_mean<-c(round(mean(support[,1]),2),round(mean(support[,2]),2),round(mean(support[,3]),2),round(mean(support[,4]),2),round(mean(support[,5]),2),round(mean(support[,6]),2),round(mean(support[,7]),2),round(mean(support[,8]),2),round((sum(support$salary=="low")+sum(support$salary=="medium")*2+sum(support$salary=="high")*3)/nrow(support),2))

technical_mean<-c(round(mean(technical[,1]),2),round(mean(technical[,2]),2),round(mean(technical[,3]),2),round(mean(technical[,4]),2),round(mean(technical[,5]),2),round(mean(technical[,6]),2),round(mean(technical[,7]),2),round(mean(technical[,8]),2),round((sum(technical$salary=="low")+sum(technical$salary=="medium")*2+sum(technical$salary=="high")*3)/nrow(technical),2))

mean_dept <- rbind(accounting_mean,hr_mean,IT_mean,management_mean,marketing_mean,product_mng_mean,RandD_mean,sales_mean,support_mean, technical_mean)

rownames(mean_dept) <- c("accounting","HR","IT", "management","marketing","product management","research and development","sales","support","technical")

colnames(mean_dept) <- c("satisfaction level","last evaluation","number of projects","average monthly hours","time spent in the company","work accident","left company","promotion during last 5 years","salary category")

iprint.df(mean_dept)
```



We calculate the standard deviation of each parameter.
```{r echo=FALSE}
accounting_sd<-c(round(sd(accounting[,1]),2),round(sd(accounting[,2]),2),round(sd(accounting[,3]),2),round(sd(accounting[,4]),2),round(sd(accounting[,5]),2),round(sd(accounting[,6]),2),round(sd(accounting[,7]),2),round(sd(accounting[,8]),2),round((sum(accounting$salary=="low")+sum(accounting$salary=="medium")*2+sum(accounting$salary=="high")*3)/nrow(accounting),2))

hr_sd<-c(round(sd(hr[,1]),2),round(sd(hr[,2]),2),round(sd(hr[,3]),2),round(sd(hr[,4]),2),round(sd(hr[,5]),2),round(sd(hr[,6]),2),round(sd(hr[,7]),2),round(sd(hr[,8]),2),round((sum(hr$salary=="low")+sum(hr$salary=="medium")*2+sum(hr$salary=="high")*3)/nrow(hr),2))

IT_sd<-c(round(sd(IT[,1]),2),round(sd(IT[,2]),2),round(sd(IT[,3]),2),round(sd(IT[,4]),2),round(sd(IT[,5]),2),round(sd(IT[,6]),2),round(sd(IT[,7]),2),round(sd(IT[,8]),2),round((sum(IT$salary=="low")+sum(IT$salary=="medium")*2+sum(IT$salary=="high")*3)/nrow(IT),2))

management_sd<-c(round(sd(management[,1]),2),round(sd(management[,2]),2),round(sd(management[,3]),2),round(sd(management[,4]),2),round(sd(management[,5]),2),round(sd(management[,6]),2),round(sd(management[,7]),2),round(sd(management[,8]),2),round((sum(management$salary=="low")+sum(management$salary=="medium")*2+sum(management$salary=="high")*3)/nrow(management),2))

marketing_sd<-c(round(sd(marketing[,1]),2),round(sd(marketing[,2]),2),round(sd(marketing[,3]),2),round(sd(marketing[,4]),2),round(sd(marketing[,5]),2),round(sd(marketing[,6]),2),round(sd(marketing[,7]),2),round(sd(marketing[,8]),2),round((sum(marketing$salary=="low")+sum(marketing$salary=="medium")*2+sum(marketing$salary=="high")*3)/nrow(marketing),2))

product_mng_sd<-c(round(sd(product_mng[,1]),2),round(sd(product_mng[,2]),2),round(sd(product_mng[,3]),2),round(sd(product_mng[,4]),2),round(sd(product_mng[,5]),2),round(sd(product_mng[,6]),2),round(sd(product_mng[,7]),2),round(sd(product_mng[,8]),2),round((sum(product_mng$salary=="low")+sum(product_mng$salary=="medium")*2+sum(product_mng$salary=="high")*3)/nrow(product_mng),2))

RandD_sd<-c(round(sd(RandD[,1]),2),round(sd(RandD[,2]),2),round(sd(RandD[,3]),2),round(sd(RandD[,4]),2),round(sd(RandD[,5]),2),round(sd(RandD[,6]),2),round(sd(RandD[,7]),2),round(sd(RandD[,8]),2),round((sum(RandD$salary=="low")+sum(RandD$salary=="medium")*2+sum(RandD$salary=="high")*3)/nrow(RandD),2))

sales_sd<-c(round(sd(sales[,1]),2),round(sd(sales[,2]),2),round(sd(sales[,3]),2),round(sd(sales[,4]),2),round(sd(sales[,5]),2),round(sd(sales[,6]),2),round(sd(sales[,7]),2),round(sd(sales[,8]),2),round((sum(sales$salary=="low")+sum(sales$salary=="medium")*2+sum(sales$salary=="high")*3)/nrow(sales),2))

support_sd<-c(round(sd(support[,1]),2),round(sd(support[,2]),2),round(sd(support[,3]),2),round(sd(support[,4]),2),round(sd(support[,5]),2),round(sd(support[,6]),2),round(sd(support[,7]),2),round(sd(support[,8]),2),round((sum(support$salary=="low")+sum(support$salary=="medium")*2+sum(support$salary=="high")*3)/nrow(support),2))

technical_sd<-c(round(sd(technical[,1]),2),round(sd(technical[,2]),2),round(sd(technical[,3]),2),round(sd(technical[,4]),2),round(sd(technical[,5]),2),round(sd(technical[,6]),2),round(sd(technical[,7]),2),round(sd(technical[,8]),2),round((sum(technical$salary=="low")+sum(technical$salary=="medium")*2+sum(technical$salary=="high")*3)/nrow(technical),2))

sd_dept <- rbind(accounting_sd,hr_sd,IT_sd,management_sd,marketing_sd,product_mng_sd,RandD_sd,sales_sd,support_sd, technical_sd)

rownames(sd_dept) <- c("accounting","HR","IT", "management","marketing","product management","research and development","sales","support","technical")

colnames(sd_dept) <- c("satisfaction level","last evaluation","number of projects","average monthly hours","time spent in the company","work accident","left company","promotion during last 5 years","salary category")

iprint.df(sd_dept)
```

We calculate the minimum value of each parameter.
```{r echo=FALSE}
accounting_min<-c(round(min(accounting[,1]),2),round(min(accounting[,2]),2),round(min(accounting[,3]),2),round(min(accounting[,4]),2),round(min(accounting[,5]),2),round(min(accounting[,6]),2),round(min(accounting[,7]),2),round(min(accounting[,8]),2),round((sum(accounting$salary=="low")+sum(accounting$salary=="medium")*2+sum(accounting$salary=="high")*3)/nrow(accounting),2))

hr_min<-c(round(min(hr[,1]),2),round(min(hr[,2]),2),round(min(hr[,3]),2),round(min(hr[,4]),2),round(min(hr[,5]),2),round(min(hr[,6]),2),round(min(hr[,7]),2),round(min(hr[,8]),2),round((sum(hr$salary=="low")+sum(hr$salary=="medium")*2+sum(hr$salary=="high")*3)/nrow(hr),2))

IT_min<-c(round(min(IT[,1]),2),round(min(IT[,2]),2),round(min(IT[,3]),2),round(min(IT[,4]),2),round(min(IT[,5]),2),round(min(IT[,6]),2),round(min(IT[,7]),2),round(min(IT[,8]),2),round((sum(IT$salary=="low")+sum(IT$salary=="medium")*2+sum(IT$salary=="high")*3)/nrow(IT),2))

management_min<-c(round(min(management[,1]),2),round(min(management[,2]),2),round(min(management[,3]),2),round(min(management[,4]),2),round(min(management[,5]),2),round(min(management[,6]),2),round(min(management[,7]),2),round(min(management[,8]),2),round((sum(management$salary=="low")+sum(management$salary=="medium")*2+sum(management$salary=="high")*3)/nrow(management),2))

marketing_min<-c(round(min(marketing[,1]),2),round(min(marketing[,2]),2),round(min(marketing[,3]),2),round(min(marketing[,4]),2),round(min(marketing[,5]),2),round(min(marketing[,6]),2),round(min(marketing[,7]),2),round(min(marketing[,8]),2),round((sum(marketing$salary=="low")+sum(marketing$salary=="medium")*2+sum(marketing$salary=="high")*3)/nrow(marketing),2))

product_mng_min<-c(round(min(product_mng[,1]),2),round(min(product_mng[,2]),2),round(min(product_mng[,3]),2),round(min(product_mng[,4]),2),round(min(product_mng[,5]),2),round(min(product_mng[,6]),2),round(min(product_mng[,7]),2),round(min(product_mng[,8]),2),round((sum(product_mng$salary=="low")+sum(product_mng$salary=="medium")*2+sum(product_mng$salary=="high")*3)/nrow(product_mng),2))

RandD_min<-c(round(min(RandD[,1]),2),round(min(RandD[,2]),2),round(min(RandD[,3]),2),round(min(RandD[,4]),2),round(min(RandD[,5]),2),round(min(RandD[,6]),2),round(min(RandD[,7]),2),round(min(RandD[,8]),2),round((sum(RandD$salary=="low")+sum(RandD$salary=="medium")*2+sum(RandD$salary=="high")*3)/nrow(RandD),2))

sales_min<-c(round(min(sales[,1]),2),round(min(sales[,2]),2),round(min(sales[,3]),2),round(min(sales[,4]),2),round(min(sales[,5]),2),round(min(sales[,6]),2),round(min(sales[,7]),2),round(min(sales[,8]),2),round((sum(sales$salary=="low")+sum(sales$salary=="medium")*2+sum(sales$salary=="high")*3)/nrow(sales),2))

support_min<-c(round(min(support[,1]),2),round(min(support[,2]),2),round(min(support[,3]),2),round(min(support[,4]),2),round(min(support[,5]),2),round(min(support[,6]),2),round(min(support[,7]),2),round(min(support[,8]),2),round((sum(support$salary=="low")+sum(support$salary=="medium")*2+sum(support$salary=="high")*3)/nrow(support),2))

technical_min<-c(round(min(technical[,1]),2),round(min(technical[,2]),2),round(min(technical[,3]),2),round(min(technical[,4]),2),round(min(technical[,5]),2),round(min(technical[,6]),2),round(min(technical[,7]),2),round(min(technical[,8]),2),round((sum(technical$salary=="low")+sum(technical$salary=="medium")*2+sum(technical$salary=="high")*3)/nrow(technical),2))

min_dept <- rbind(accounting_min,hr_min,IT_min,management_min,marketing_min,product_mng_min,RandD_min,sales_min,support_min, technical_min)

rownames(min_dept) <- c("accounting","HR","IT", "management","marketing","product management","research and development","sales","support","technical")

colnames(min_dept) <- c("satisfaction level","last evaluation","number of projects","average monthly hours","time spent in the company","work accident","left company","promotion during last 5 years","salary category")

iprint.df(min_dept)
```

We calculate the maxiimum value of each parameter.
```{r echo=FALSE}
accounting_max<-c(round(max(accounting[,1]),2),round(max(accounting[,2]),2),round(max(accounting[,3]),2),round(max(accounting[,4]),2),round(max(accounting[,5]),2),round(max(accounting[,6]),2),round(max(accounting[,7]),2),round(max(accounting[,8]),2),round((sum(accounting$salary=="low")+sum(accounting$salary=="medium")*2+sum(accounting$salary=="high")*3)/nrow(accounting),2))

hr_max<-c(round(max(hr[,1]),2),round(max(hr[,2]),2),round(max(hr[,3]),2),round(max(hr[,4]),2),round(max(hr[,5]),2),round(max(hr[,6]),2),round(max(hr[,7]),2),round(max(hr[,8]),2),round((sum(hr$salary=="low")+sum(hr$salary=="medium")*2+sum(hr$salary=="high")*3)/nrow(hr),2))

IT_max<-c(round(max(IT[,1]),2),round(max(IT[,2]),2),round(max(IT[,3]),2),round(max(IT[,4]),2),round(max(IT[,5]),2),round(max(IT[,6]),2),round(max(IT[,7]),2),round(max(IT[,8]),2),round((sum(IT$salary=="low")+sum(IT$salary=="medium")*2+sum(IT$salary=="high")*3)/nrow(IT),2))

management_max<-c(round(max(management[,1]),2),round(max(management[,2]),2),round(max(management[,3]),2),round(max(management[,4]),2),round(max(management[,5]),2),round(max(management[,6]),2),round(max(management[,7]),2),round(max(management[,8]),2),round((sum(management$salary=="low")+sum(management$salary=="medium")*2+sum(management$salary=="high")*3)/nrow(management),2))

marketing_max<-c(round(max(marketing[,1]),2),round(max(marketing[,2]),2),round(max(marketing[,3]),2),round(max(marketing[,4]),2),round(max(marketing[,5]),2),round(max(marketing[,6]),2),round(max(marketing[,7]),2),round(max(marketing[,8]),2),round((sum(marketing$salary=="low")+sum(marketing$salary=="medium")*2+sum(marketing$salary=="high")*3)/nrow(marketing),2))

product_mng_max<-c(round(max(product_mng[,1]),2),round(max(product_mng[,2]),2),round(max(product_mng[,3]),2),round(max(product_mng[,4]),2),round(max(product_mng[,5]),2),round(max(product_mng[,6]),2),round(max(product_mng[,7]),2),round(max(product_mng[,8]),2),round((sum(product_mng$salary=="low")+sum(product_mng$salary=="medium")*2+sum(product_mng$salary=="high")*3)/nrow(product_mng),2))

RandD_max<-c(round(max(RandD[,1]),2),round(max(RandD[,2]),2),round(max(RandD[,3]),2),round(max(RandD[,4]),2),round(max(RandD[,5]),2),round(max(RandD[,6]),2),round(max(RandD[,7]),2),round(max(RandD[,8]),2),round((sum(RandD$salary=="low")+sum(RandD$salary=="medium")*2+sum(RandD$salary=="high")*3)/nrow(RandD),2))

sales_max<-c(round(max(sales[,1]),2),round(max(sales[,2]),2),round(max(sales[,3]),2),round(max(sales[,4]),2),round(max(sales[,5]),2),round(max(sales[,6]),2),round(max(sales[,7]),2),round(max(sales[,8]),2),round((sum(sales$salary=="low")+sum(sales$salary=="medium")*2+sum(sales$salary=="high")*3)/nrow(sales),2))

support_max<-c(round(max(support[,1]),2),round(max(support[,2]),2),round(max(support[,3]),2),round(max(support[,4]),2),round(max(support[,5]),2),round(max(support[,6]),2),round(max(support[,7]),2),round(max(support[,8]),2),round((sum(support$salary=="low")+sum(support$salary=="medium")*2+sum(support$salary=="high")*3)/nrow(support),2))

technical_max<-c(round(max(technical[,1]),2),round(max(technical[,2]),2),round(max(technical[,3]),2),round(max(technical[,4]),2),round(max(technical[,5]),2),round(max(technical[,6]),2),round(max(technical[,7]),2),round(max(technical[,8]),2),round((sum(technical$salary=="low")+sum(technical$salary=="medium")*2+sum(technical$salary=="high")*3)/nrow(technical),2))

max_dept <- rbind(accounting_max,hr_max,IT_max,management_max,marketing_max,product_mng_max,RandD_max,sales_max,support_max, technical_max)

rownames(max_dept) <- c("accounting","HR","IT", "management","marketing","product management","research and development","sales","support","technical")

colnames(max_dept) <- c("satisfaction level","last evaluation","number of projects","average monthly hours","time spent in the company","work accident","left company","promotion during last 5 years","salary category")

iprint.df(max_dept)
```

# Factor Analysis

```{r setupfactor, echo=TRUE, tidy=TRUE}
# Please ENTER then original raw attributes to use. 
# Please use numbers, not column names, e.g. c(1:5, 7, 8) uses columns 1,2,3,4,5,7,8
factor_attributes_used = c(1:10)

# Please ENTER the selection criterions for the factors to use. 
# Choices: "eigenvalue", "variance", "manual"
factor_selectionciterion = "manual"

# Please ENTER the desired minumum variance explained 
# (Only used in case "variance" is the factor selection criterion used). 
minimum_variance_explained = 65  # between 1 and 100

# Please ENTER the number of factors to use 
# (Only used in case "manual" is the factor selection criterion used).
manual_numb_factors_used = 3

# Please ENTER the rotation eventually used (e.g. "none", "varimax", "quatimax", "promax", "oblimin", "simplimax", and "cluster" - see help(principal)). Default is "varimax"
rotation_used = "varimax"

```


First, we apply the factors and scale the data 

```{r echo=FALSE}
factor_attributes_used <- intersect(factor_attributes_used, 1:ncol(ProjectData))
ProjectDataFactor <- ProjectData[,factor_attributes_used]
ProjectDataFactor <- ProjectData <- data.matrix(ProjectDataFactor)

ProjectDatafactor_scaled = apply(ProjectDataFactor, 2, function(r) {
    if (sd(r) != 0) 
        res = (r - mean(r))/sd(r) else res = 0 * r
    res
})
ProjectDataFactor <- ProjectDatafactor_scaled
```

# Visual exploration of the data

```{r echo=FALSE}
rownames(ProjectDataFactor) <- paste0("Obs.", sprintf("%02i", 1:nrow(ProjectDataFactor)))
iprint.df(t(head(round(ProjectDataFactor, 2), max_data_report)))
```

The data we use here have the following descriptive statistics: 

```{r echo=FALSE}
iprint.df(round(my_summary(ProjectDataFactor), 2))
```

## Check Correlations

This is the correlation matrix of the customer responses to the `r ncol(ProjectDataFactor)` attitude questions - which are the only questions that we will use for the segmentation (see the case):

```{r echo=FALSE}
thecor = round(cor(ProjectDataFactor),2)
iprint.df(round(thecor,2), scale=TRUE)
```


## Choose number of factors - Principal Factor analysis

```{r echo=FALSE}
# Here is how the `principal` function is used 
UnRotated_Results<-principal(ProjectDataFactor, nfactors=ncol(ProjectDataFactor), rotate="none",score=TRUE)
UnRotated_Factors<-round(UnRotated_Results$loadings,2)
UnRotated_Factors<-as.data.frame(unclass(UnRotated_Factors))
colnames(UnRotated_Factors)<-paste("Comp",1:ncol(UnRotated_Factors),sep="")
```

```{r echo=FALSE}
# Here is how we use the `PCA` function 
Variance_Explained_Table_results<-PCA(ProjectDataFactor, graph=FALSE)
Variance_Explained_Table<-Variance_Explained_Table_results$eig
Variance_Explained_Table_copy<-Variance_Explained_Table

rownames(Variance_Explained_Table) <- paste("Component", 1:nrow(Variance_Explained_Table), sep=" ")
colnames(Variance_Explained_Table) <- c("Eigenvalue", "Pct of explained variance", "Cumulative pct of explained variance")
```

Let's look at the **variance explained** as well as the **eigenvalues** :

```{r echo=FALSE}
iprint.df(round(Variance_Explained_Table, 2))
```

```{r echo=FALSE}
eigenvalues  <- Variance_Explained_Table[, "Eigenvalue"]
df           <- cbind(as.data.frame(eigenvalues), c(1:length(eigenvalues)), rep(1, length(eigenvalues)))
colnames(df) <- c("eigenvalues", "components", "abline")
iplot.df(melt(df, id="components"))
```

# Interpret the factors

Let's now see how the "top factors" look like. 

```{r echo=FALSE}
if (factor_selectionciterion == "eigenvalue")
  factors_selected = sum(Variance_Explained_Table_copy[,1] >= 1)
if (factor_selectionciterion == "variance")
  factors_selected = 1:head(which(Variance_Explained_Table_copy[,"cumulative percentage of variance"]>= minimum_variance_explained),1)
if (factor_selectionciterion == "manual")
  factors_selected = manual_numb_factors_used
```

# Rotation applied on factors

```{r echo=FALSE}
Rotated_Results<-principal(ProjectDataFactor, nfactors=max(factors_selected), rotate=rotation_used,score=TRUE)
Rotated_Factors<-round(Rotated_Results$loadings,2)
Rotated_Factors<-as.data.frame(unclass(Rotated_Factors))
colnames(Rotated_Factors)<-paste("Comp.",1:ncol(Rotated_Factors),sep="")

sorted_rows <- sort(Rotated_Factors[,1], decreasing = TRUE, index.return = TRUE)$ix
Rotated_Factors <- Rotated_Factors[sorted_rows,]

iprint.df(Rotated_Factors, scale=TRUE)
```

To better visualize and interpret the factors we often "suppress" loadings with small values, e.g. with absolute values smaller than 0.5. In this case our factors look as follows after suppressing the small numbers:

```{r echo=FALSE}
Rotated_Factors_thres <- Rotated_Factors
Rotated_Factors_thres[abs(Rotated_Factors_thres) < MIN_VALUE]<-NA
colnames(Rotated_Factors_thres)<- colnames(Rotated_Factors)
rownames(Rotated_Factors_thres)<- rownames(Rotated_Factors)

iprint.df(Rotated_Factors_thres, scale=TRUE)
```

# Save factor scores 

We can now either replace all initial variables used in this part with the factors scores or just select one of the initial variables for each of the selected factors in order to represent that factor. Here is how the factor scores  are for the first few respondents:

```{r echo=FALSE}
NEW_ProjectData <- round(Rotated_Results$scores[,1:factors_selected,drop=F],2)
colnames(NEW_ProjectData)<-paste("DV (Factor)",1:ncol(NEW_ProjectData),sep=" ")

iprint.df(t(head(NEW_ProjectData, 10)), scale=TRUE)
```


## Answering the Business Questions:

*1. Why are employees leaving?*

To sum up, we can say employees who left:
- were all good performers
- used to work for long hours in the office
- were most likely not promoted within the last few years
- were unsatisfied employees have been members of just two projects or strictly more than five


*2. What are the three most important factors for people leaving as compared to staying?*

*a) Are we overworking the top performers?*

This is a possible conclusion, however this would indicate that the employees that left would have a higher evaluation than those who stayed which and work longer hours than those who stayed. However, according to the data, even though those who left worked longer hours, they did not have a higher or lower evaluation.

*b) Is money most important? Is satisfaction most important?*

The data shows that satisfaction may be an important factor as those who left had a lower level of satisfaction with their job than those who stayed. 
The raw data captured only split the salary into high, medium and low categories. Withouot more speicifc data it is difficult to draw a conclusion if salary plays a significant part in employe eretention 

*c) Department vs. Global Trends - do we have bad managers?*

We don't see any significant differences across the departments. 

*3. How can we keep employees using the information that we derive from the data? Can we identify people who are leaving *

We have derived that the employees that left were good performers, were possibly overworked as they worked more long hours in the office, had not neccessarily received a promotion within the last few years and have been members of either two long projects or more than five (many short projects)

Therefore each of these factors can be used to put a "red flag" on anyone who displays these characteristics to ensure that they are satisfied and catch the problems early.

*4. Can we identify loyal employees?*

According to the statistics loyal employees are those that would stay for longer despite working long hours, or have a recent lower level of satisfaction.

*5. Turnover per department?* 

The turnover per department is as follows

```{r echo=FALSE}
iprint.df(mean_dept[,7])
```
